A Formal Model of Checked C

### 2023-01-18 sandp23/background.tex
update section 1 and 2.

\section{Background and Motivation}
%\section{Overview and Transcendence}\label{sec:overview}
%This section describes \checkedc{} and new features \systemname{} provides.
This section presents a brief background of~\checkedc and motivation for~\systemname{}.

\subsection{\checkedc}
\label{subsec:checkedc}
% 
\checkedc{} development began in 2015 by Microsoft Research, but it was forked
in late 2021 and is now actively managed by the Secure Software
Development Project (SSDP). Details can be found in a prior
overview~\cite{Elliott2018} or the formalism~\cite{li22checkedc}.
%

\liyi{Do we need to spend more than half page introducing checked-c? OR just introduce it briefly like what we had in the formalism paper? }

@@

  &
    $[x, y)    $\\  
\end{tabular}
\end{footnotesize}
The bounds can be declared for $\ntarr$ as well, but the memory range can extend further to the right,
until a~\code{NULL} terminator is reached (\ie the~\code{NULL} is not within the bounds).

\liyi{can we reuse the old text for the spatial safety? }

\noindent
\myparagraph{Ensuring Spatial Memory Safety}
The \checkedc compiler will instrument loads and stores of checked
pointers to confirm the pointer is non-null, and additionally for $\arr$ and $\ntarr$ pointers the access is within
the specified bounds.
For instance, in the code \inlinecode{if (n>0) a[n-1] =} $...$ the write
is via address $\alpha = \inlinecode{a + sizeof(int)}\!\times\!\code{(n-1)}$. 
If the bounds of \code{a} are \code{count(u)}, the
inserted check will confirm that prior to dereference
$\inlinecode{a} \leq \alpha < \inlinecode{a + sizeof(int)} \!\times\!
\inlinecode{u}$.
Failed checks throw an exception.
Oftentimes, inserted checks can be optimized away by LLVM resulting in almost no runtime overhead~\cite{duanrefactoring}.

@@

\noindent
\myparagraph{Backward Compatibility}
\checkedc is backward compatible with legacy C in the sense that all legacy code will type-check and compile.
However, the compiler adds the aforementioned spatial safety checks to only checked pointers.
The spatial safety guarantee is partial when the code is not fully ported.
Specifically, only code that appears in \emph{checked
  code regions}, which we call \emph{checked code}, is guaranteed to be spatially safe.
Checked regions can be designated at the level of files, functions, or individual code blocks using the
\code{checked} keyword.\footnote{You can also designate~\emph{unchecked} regions  within checked ones.}
 Within checked regions, both legacy pointers and
certain unsafe idioms (\eg \emph{variadic} function calls) are disallowed.

\liyi{If we are not going to talk about converting C to \checkedc, can we just move this to future work? }

@@

\liyi{Does the program too small? Will reviewer complain about the program size? }
\subsection{No Safety Against Unchecked Code}
\label{subsec:nosafetyagsintuncheckedcode}
The spatial safety guarantee~\checkedc provides applies to completely converted programs,~\ie programs that uses~\emph{only} checked types and no regular pointer types.
A partially annotated program can still enjoy spatial safety but only on those code regions using only Checked pointers (\ie checked or safe regions). For instance, in the below example, there can be no spatial safety violations in the function~\code{func} as it uses only checked pointers.
However, the other unconverted code regions (or unsafe regions) can affect pointers in safe regions and violate certain assumptions leading to vulnerabilities, as demonstrated by cross-language attacks~\cite{mergendahlcross}.

%The backward compatibility of~\checkedc helps a partially annotated program to enjoy spatial memory safety on those regions using only Checked pointers (\ie checked or safe regions).
%But, these unconverted code regions (or unsafe regions) can affect pointers in safe regions and violate certain assumptions leading to vulnerabilities, as demonstrated by cross-language attacks~\cite{mergendahlcross}. \lc{I'm confused, does the partially checked program "enjoy spatial memory safety" or not?}
Although the blameless proof exists~\cite{ruef2019achieving, li22checkedc} for~\checkedc, it does not state that spatial safety violations cannot happen in Checked regions but rather states that Checked regions~\emph{cannot be blamed for any spatial safety violations}.
Consider the following example:
\begin{minted}[xleftmargin=25pt, mathescape, linenos, escapeinside=||, fontsize=\footnotesize{}]{c}
// Checked code
int func(array_ptr<char> p : count(5)) {
|\textcolor{red}{\faChainBroken}|..p[4]..
}
// unchecked code
...
str = "he";
...
func(|\textcolor{red}{\faBug}|assume_bounds_cast<char>(str, 5)); 
\end{minted}
Here, the checked function~\inlinecode{func} expected a pointer to a buffer of five elements, but unchecked code violated it and invoked the function (Line 9) with a buffer of 2 elements.
This results in a spatial safety violation (\textcolor{red}{\faChainBroken}) in the Checked region, but of course, the blame or the root cause is in the unchecked region (\textcolor{red}{\faBug}).
Furthermore, since checked and unchecked regions execute in the same address space, spatial memory corruptions (\eg buffer overflow) in unchecked regions can take down the complete program despite having checked regions.
We need~\emph{an isolation mechanism to ensure that code executed as part of unchecked regions does not violate the safety guarantees in checked regions}.

>>>>

\section{Background and Motivation}
%\section{Overview and Transcendence}\label{sec:overview}
%This section describes \checkedc{} and new features \systemname{} provides.
Here, we brief \checkedc and the motivation for~\systemname{}.

\subsection{\checkedc}
\label{subsec:checkedc}
% 
\checkedc{} development began in 2015 by Microsoft Research, but it was forked
in late 2021 and is now actively managed by the Secure Software
Development Project (SSDP). Details can be found in a prior
overview~\cite{Elliott2018} and the formalism~\cite{li22checkedc}.

@@

  &
    $[x, y)    $\\  
\end{tabular}
\end{footnotesize}
The bounds can be declared for $\ntarr$ as well, but the memory range can extend further to the right,
until a~\code{NULL} terminator is reached (\ie \code{NULL} is not within the bounds).

\noindent
\myparagraph{Ensuring Spatial Memory Safety}
The \checkedc compiler instruments loads and stores of checked
pointers to confirm the pointer is non-null, and additionally the access to $\arr$ and $\ntarr$ pointers is within
their specified bounds.
For example, in the code \inlinecode{if (n>0) a[n-1] =} $...$ the write
is via address $\alpha = \inlinecode{a + sizeof(int)}\!\times\!\code{(n-1)}$. 
If the bounds of \code{a} are \code{count(u)}, the
inserted check confirms
$\inlinecode{a} \leq \alpha < \inlinecode{a + sizeof(int)} \!\times\!
\inlinecode{u}$ prior to dereference.
Failed checks throw an exception.
Oftentimes, inserted checks can be optimized away by LLVM resulting in almost no runtime overhead~\cite{duanrefactoring}.

@@

\noindent
\myparagraph{Backward Compatibility}
\checkedc is backward compatible with legacy C as all legacy code will type-check and compile.
However, the compiler adds the aforementioned spatial safety checks to only checked pointers.
The spatial safety guarantee is partial when the code is not fully ported.
Specifically, only code that appears in \emph{checked
  code regions} (\cregion), is guaranteed to be spatially safe.
$\cmode$ regions can be designated at the level of files, functions, or individual code blocks using the
\code{checked} keyword.\footnote{You can also designate~\emph{unchecked} regions (\ucregion) within checked ones.}
 Within $\cmode$ regions, both legacy pointers and
certain unsafe idioms (\eg \emph{variadic} function calls) are disallowed.

@@

\subsection{No Safety Against $\umode$ Regions}
\label{subsec:nosafetyagsintuncheckedcode}
\checkedc provides spatial safety guarantees for completely converted programs,~\ie programs that uses~\emph{only} checked types and no regular pointer types.
A partially annotated program can still enjoy spatial safety only if checked pointers do not communicate with any unchecked ones. For instance, in the example below, there are no spatial safety violations in the function~\code{func} as it uses only checked pointers.
However, the other unconverted code regions (or unsafe regions) can affect pointers in safe regions and violate certain assumptions leading to vulnerabilities, as demonstrated by cross-language attacks~\cite{mergendahlcross}.

%The backward compatibility of~\checkedc helps a partially annotated program to enjoy spatial memory safety on those regions using only Checked pointers (\ie checked or safe regions).
%But, these unconverted code regions (or unsafe regions) can affect pointers in safe regions and violate certain assumptions leading to vulnerabilities, as demonstrated by cross-language attacks~\cite{mergendahlcross}. \lc{I'm confused, does the partially checked program "enjoy spatial memory safety" or not?}
Although the blameless proof exists~\cite{ruef2019achieving, li22checkedc} for~\checkedc, it does not state that spatial safety violations cannot happen in $\cmode$ regions but rather states that $\cmode$ regions~\emph{cannot be blamed for any spatial safety violations}.
Consider the following example:
\begin{minted}[xleftmargin=25pt, mathescape, linenos, escapeinside=||, fontsize=\footnotesize{}]{c}
// c region code
int func(array_ptr<char> p : count(5)) {
|\textcolor{red}{\faChainBroken}|..p[4]..
}
// u region code
...
str = "he";
...
func(|\textcolor{red}{\faBug}|assume_bounds_cast<char>(str, 5)); 
\end{minted}
Here, the $\cmode$ region function~\inlinecode{func} expects a pointer to a buffer of five elements, but the $\umode$ region code
invokes the function (Line 9) with a buffer of 2 elements.
This results in a spatial safety violation (\textcolor{red}{\faChainBroken}) in the $\cmode$ region, but of course, the blame or the root cause is in the $\umode$ region (\textcolor{red}{\faBug}).
Furthermore, since $\cmode$ and $\umode$ regions execute in the same address space, spatial memory corruptions (\eg buffer overflow) in $\umode$ regions can take down the complete program despite having $\cmode$ regions.
We need~\emph{an isolation mechanism to ensure that code executed as part of $\umode$  regions does not violate the safety guarantees in $\cmode$ regions}.
